Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-14 paper/manuscript.Rmd
minor edits

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitat. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
translocations and introductions throughout the province [@Bergerud1989].
Typical caribou predators such as black bears (*Ursus americanus*) and wolves
(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
(*Canis latrans*), which predate calves but it remains unclear what threat they
pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
we propose that forage limitation is more likely to drive movement between
islands than predator avoidance. For several decades after their introduction to
Fogo Island, the caribou population did not exceed ~100 individuals
[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
During the 1990s, population density reached ~300 animals and anecdotal evidence
suggests caribou were first seen on nearby islands during this same period.
Increased competition for resources on Fogo due to this increase in density
could have precipitated movement to other islands.

Forage limitation and over-grazing is a major concern for caribou populations
and can lead to reduced female reproductive success [@Schaefer2016]. Forage
scarcity has been proposed as a potential reason for caribou moving between
arctic islands on the sea-ice in winter [@Miller1977]. For the Fogo Island
population, as the population grew and predation pressure was insufficient to
maintain population density, forage depletion could lead caribou to make
similar, but riskier and more costly, movements between islands during the
ice-free season. Given the small size of other islands near Fogo, over-grazing
by newly arrived caribou could rapidly deplete forage on an annual cycle
[@Bergerud1990], leading to more frequent movements back and forth between
islands rather than permanent relocation.

>>>>

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitats and foraging opportunities. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
translocations and introductions throughout the province [@Bergerud1989].
Typical caribou predators such as black bears (*Ursus americanus*) and wolves
(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
(*Canis latrans*), which predate calves but it remains unclear what threat they
pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
we propose that forage limitation is more likely to drive movement between
islands than predator avoidance. For several decades after their introduction to
Fogo Island, the caribou population did not exceed ~100 individuals
[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
During the 1990s, population density reached ~300 animals and anecdotal evidence
suggests caribou were first seen on nearby islands during this same period.
Increased competition for resources on Fogo Island due to this increase in density
could have precipitated movement to other islands.

Forage limitation and over-grazing is a major concern for caribou populations
and can lead to reduced female reproductive success [@Schaefer2016]. Forage
scarcity has been proposed as a potential reason for caribou moving between
arctic islands on the sea-ice in winter [@Miller1977]. For the Fogo Island
population, as the population grew and predation pressure was insufficient to
maintain population density, forage depletion could lead caribou to make
similar, but riskier and more costly, movements between islands during the
ice-free season. Given the small size of other islands near Fogo (Figure 2), over-grazing
by newly arrived caribou could rapidly deplete forage on an annual cycle
[@Bergerud1990], leading to more frequent movements back and forth between
islands rather than permanent relocation.
